---
title: "Marine boundaries"
author: "James Goldie, 360info"
---

# Introduction

```{r setup}
library(tidyverse)
library(sf)
library(countrycode)
library(themes360info)
library(ggtext)
library(here)
```
Exclusive Economic Zones (EEZ) mark out the areas in which sovereign states can
extract and use natural resources to the exclusion of others. Many EEZs really
are exclusive to one state, but some EEZs are jointly managed by two or more
sovereign states, while others are overlapping claims.

Let's load in the boundaries of the _deep sea_ regions of these EEZ. This has
been calculated in QGIS using two data sources:

1. **EEZ boundaries:** Flanders Marine Institute (2019). Maritime Boundaries
Geodatabase: Maritime Boundaries and Exclusive Economic Zones (200NM),
version 11. Available online at https://www.marineregions.org/.
https://doi.org/10.14284/386
2. **Bathymetry (elevation):** GEBCO Compilation Group (2021) GEBCO 2021 Grid
https://doi.org/10.5285/c6612cbe-50b3-0cff-e053-6c86abc09f8f

```{r import}

# first we need to grab our qgis files from the 360 google drive
download.file(
  paste0(
    "https://drive.google.com/uc?export=download&confirm=t&",
    "id=1NVJeEvIzeGcXCgRCs3IlXXD0xXVfvWfI"),
  destfile = here("data", "regions-below-200m.gpkg"))
download.file(
  paste0(
    "https://drive.google.com/uc?export=download&confirm=t&",
    "id=1Q8P4qY9bpJGSJ6zhB3SG8UrkHEHkaekp"),
  destfile = here("data", "full-eez.gpkg"))
download.file(
  paste0(
    "https://drive.google.com/uc?export=download&confirm=t&",
    "id=1dvpLpHRcCdLVF5otz6vrXMgoYYaKjZjy"),
  destfile = here("data", "deepsea-eez-boundaries.gpkg"))

eez_deep <-
  read_sf(here("data", "deepsea-eez-regions.gpkg")) %>%
  select(
    id_num = MRGID,
    zone_name = eez_v11_GEONAME,
    zone_type = eez_v11_POL_TYPE,
    # the whole area of the eez (including the shallows aove 200m, which are not
    # part of the geometry of this file)
    total_area = eez_v11_AREA_KM2,
    # the area of the deep sea - sub 200m - part of the eez, which this geometry shows
    deepsea_area_km2 = deepsea_area_km2,
    claim1_territory_name = eez_v11_TERRITORY1,
    claim1_territory_iso3 = eez_v11_ISO_TER1,
    claim1_sov_name       = eez_v11_SOVEREIGN1,
    claim1_sov_iso3       = eez_v11_ISO_SOV1,
    claim2_territory_name = eez_v11_TERRITORY2,
    claim2_territory_iso3 = eez_v11_ISO_TER2,
    claim2_sov_name       = eez_v11_SOVEREIGN2,
    claim2_sov_iso3       = eez_v11_ISO_SOV2,
    claim3_territory_name = eez_v11_TERRITORY3,
    claim3_territory_iso3 = eez_v11_ISO_TER3,
    claim3_sov_name       = eez_v11_SOVEREIGN3,
    claim3_sov_iso3       = eez_v11_ISO_SOV3,
    geom) %>%
  mutate(frac_deepsea = deepsea_area_km2 / total_area)
```

Now we can answer some questions!

# How much of the deep sea is claimed?

How much deep sea is there in total, and how much of it is under _any_ claim?

```{r allclaims}
# let's load a separate file of all the deep sea regins on earth to get that
# area
depth_classes <-
  read_sf(here("data", "regions-below-200m.gpkg")) %>%
  st_transform(crs = 8857) %>%
  mutate(area_km2 = as.numeric(st_area(.)) / 1e6)
deepsea_area <-
  depth_classes %>%
  filter(description == "Sub 200m Depth") %>%
  pull(area_km2)

  # note that alt

# now sum up the claim areas (with and without antarctica;s eez)
all_claims <-
  eez_deep %>%
  pull(deepsea_area_km2) %>%
  sum(na.rm = TRUE)
all_claims_less_antarctica <-
  eez_deep %>%
  filter(!str_detect(zone_name, "Antarctic")) %>%
  pull(deepsea_area_km2) %>%
  sum(na.rm = TRUE)

# area labeller
label_area <- scales::label_number_si(" square kilometres", accuracy = 1)
```

So the deep sea comprises `r label_area(deepsea_area)`, of which
`r scales::percent(all_claims / as.numeric(deepsea_area))` is claimed by one or
more states as an EEZ.

There is some complicating detail here. Parts of the Arctic and Antarctic are
covered by sea ice but have deep sea underneath, which is included here.

The Antarctic also has an EEZ extending out from its boundary (even though
there's deep sea behind that boundary). If you remove the Antarctic EEZ but
still include the deep sea behind its boundary, the total fraction of the deep
sea claimed falls to
`r scales::percent(all_claims_less_antarctica / as.numeric(deepsea_area))`.

# Total EEZ area by sovereign state

To work this out, let's tidy this dataset so that there's one row per _claim_ instead of one row per EEZ:

```{r tidyclaims}

# let's isolate the unique deepsea info and merge it into the full eez shapes
# (which include shallow-only eezs)
eez_deep %>%
  as_tibble() %>%
  select(id_num, deepsea_area_km2, frac_deepsea) ->
eez_deep_info

eez_full <-
  read_sf(here("data", "full-eez-regions.gpkg")) %>%
  as_tibble() %>%
  select(
    id_num = MRGID,
    zone_name = GEONAME,
    zone_type = POL_TYPE,
    total_area = AREA_KM2,
    claim1_territory_name = TERRITORY1,
    claim1_territory_iso3 = ISO_TER1,
    claim1_sov_name       = SOVEREIGN1,
    claim1_sov_iso3       = ISO_SOV1,
    claim2_territory_name = TERRITORY2,
    claim2_territory_iso3 = ISO_TER2,
    claim2_sov_name       = SOVEREIGN2,
    claim2_sov_iso3       = ISO_SOV2,
    claim3_territory_name = TERRITORY3,
    claim3_territory_iso3 = ISO_TER3,
    claim3_sov_name       = SOVEREIGN3,
    claim3_sov_iso3       = ISO_SOV3,
    geom)

# join the two datasets together
eez_full %>%
  left_join(eez_deep_info, by = "id_num") %>%
  replace_na(list(deepsea_area_km2 = 0, frac_deepsea = 0)) ->
eez_joined

# now let's lengthen this to work out which areas _any_ sovereign state has
# access to
eez_joined %>%
  select(-geom) %>%
  # lengthen claim number and territory/sovereign
  pivot_longer(
    starts_with("claim"),
    names_to = c("claim_num", "party", ".value"), names_sep = "_") %>%
  # now go wider again
  pivot_wider(names_from = "party", values_from = c("name", "iso3")) %>%
  select(-claim_num) %>%
  # now drop the rows where there isn't a second or third claim
  drop_na(starts_with("_territory"), starts_with("_sov")) %>%
  write_csv(here("data", "eez-claims-tidy.csv")) ->
eez_claims

eez_claims
```

Now let's add up the area each sovereign state controls by zone type:

```{r areabysov}
eez_claims %>%
  group_by(name_sov, zone_type) %>%
  summarise(
    deep = sum(deepsea_area_km2, na.rm = TRUE),
    shallow = sum(total_area, na.rm = TRUE) - deep) %>%
  pivot_longer(where(is.double), names_to = "depth_class", values_to = "area") ->
eez_totals_by_sov

# we want to order the bars by the size of the deep sea area each sovereign
# controls, regardless of the exclusivity. let's calculate that separately
# and merge it back in rather than relying on factor levels
eez_totals_by_sov %>%
  filter(depth_class == "deep") %>%
  group_by(name_sov) %>%
  summarise(deep_area = sum(area, na.rm = TRUE)) %>%
  arrange(desc(deep_area)) %>%
  mutate(
    area_rank = 1:n(),
    name_sov = fct_reorder(name_sov, deep_area),
    name_lump = fct_lump_n(name_sov, n = 20, w = deep_area,
      other_level = "Other countries")) ->
sov_order

eez_totals_by_sov %>%
  left_join(sov_order, by = "name_sov") %>%
  select(-deep_area) ->
ordered_eez_totals_by_sov

# plot colours
bar_colours <- c(
  "Overlapping claim" = mnsl2hex("5YR 7/12"),
  "Joint regime" = mnsl2hex("7.5G 8/10"),
  "Exclusive" = mnsl2hex("5PB 2/6")
)

ordered_eez_totals_by_sov %>%
  mutate(
    zone_type = factor(
      if_else(zone_type == "200NM", "Exclusive", zone_type),
      levels = c("Exclusive", "Joint regime", "Overlapping claim"))) %>%
  filter(depth_class == "deep") %>%
  {
    ggplot(.) +
      aes(x = name_lump, y = area, fill = zone_type, colour = zone_type) +
      geom_col(position = position_stack(reverse = TRUE)) +
      coord_flip() +
      annotate(geom = "richtext",
        x = 6, y = 0.75e7,
        fill = NA, label.colour = NA, size = 6, hjust = "left",
        family = "Body 360info",
        label = paste(
          "**Where's the boundary?**",
          "These areas represent a state's",
          "Exclusive Economic Zone (EEZ).",
          "<br>",
          "EEZs typically extend 200 nautical miles",
          "from a country, except where another",
          "country is closer.",
          sep = "<br>"),
        ) +
      scale_fill_manual(values = bar_colours,
        guide = guide_legend(reverse = TRUE)) +
      scale_colour_manual(values = bar_colours,
        guide = guide_legend(reverse = TRUE)) +
      scale_y_continuous(
        position = "right",
        labels = scales::label_number_si(unit = "")) +
      labs(
        x = "", y = "Area (sq. km)",
        fill = "", colour = "",
        title = toupper("Access to the deep sea"),
        subtitle = toupper(
          "Area of deep sea within national maritime boundaries"),
        caption = paste(
          "**CHART:** James Goldie, 360info",
          "**SOURCES**: Flanders Marine Institute, GEBCO",
          sep = "<br>")
      ) +
      theme_360() +
      theme(
        legend.direction = "horizontal",
        legend.position = "top",
        plot.subtitle = element_text(margin = margin(0, 0, 10, 0, "pt")),
        legend.box.spacing = unit(5, "pt"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.x.top = element_text(margin = margin(0, 0, 20, 0, "pt")))
  } %>%
  print() %>%
  save_360plot(here("marine-boundaries", "out", "marine-boundaries2.png"),
    shape = "phone-portrait")
```
